Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-12-21 journal paper/content/related_work.tex
Some edits after Guy's pass, removing some color-edits.

In \citet{keller2003price}, the competition is entirely on prices, rather than product quality. This distinction leads to several important differences. First, the data externality -- that each customer brings a new data point to a firm if and only if he chooses this firm -- is absolutely crucial to our setting, and absent in theirs. Second, the inputs to agents' decision rule are prices, rather than the quality of the chosen alternatives; the former is directly controlled by the firms, whereas the latter is not known a priori. Third, the goal of exploration is to learn the agents' decision rule, \gaedit{not the best product alternative}.

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, exogenously determined, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different \gaedit{than in our setting}. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms', so that when one compares competition to the planner, one can only compare outcomes, but not the algorithms.\footnote{The planner faces a two-armed bandit problem, because the planner can directly choose a firm for each agent.}
% firm, because payments cancel out in social welfare).}

>>>>

In \citet{keller2003price}, the competition is entirely on prices, rather than product quality. This distinction leads to several important differences. First, the data externality -- that each customer brings a new data point to a firm if and only if he chooses this firm -- is absolutely crucial to our setting, and absent in theirs. Second, the inputs to agents' decision rule are prices, rather than the quality of the chosen alternatives; the former is directly controlled by the firms, whereas the latter is not known a priori. Third, the goal of exploration is to learn the agents' decision rule, not the best product alternative.

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, exogenously determined, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different than in our setting. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms', so that when one compares competition to the planner, one can only compare outcomes, but not the algorithms.\footnote{The planner faces a two-armed bandit problem, because the planner can directly choose a firm for each agent.}
% firm, because payments cancel out in social welfare).}

### 2020-12-21 journal paper/content/sec-intro.tex
Some edits after Guy's pass, removing some color-edits.

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule \gaedit{which depends on the principal's exploration strategy}. The principal selects an action which affects the quality of service provided to this agent (\eg a list of web search results). The agent experiences this action \gaedit{and the resulting utility from this action is observed by the principal.} \gadelete{and reports a reward.} Each principal only observes its own users. Principals commit to their strategies in advance, so as to maximize their market share.

@@

\textbf{(iii)} 
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance \gaedit{in a standalone bandit problem, \ie without competition, and} we find that the most natural performance measure does not explain this phenomenon. We suggest a new, more nuanced performance measure that does.

>>>>

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule \asedit{(which may depend on the principal's exploration strategy, either directly or indirectly)}. The principal selects an action which affects the quality of service provided to this agent (\eg a list of web search results). The agent experiences this action \gaedit{and the resulting utility from this action is observed by the principal.} \gadelete{and reports a reward.} Each principal only observes its own users. Principals commit to their strategies in advance, so as to maximize their market share.

@@

\textbf{(iii)}
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance in a standalone bandit problem, \ie without competition, and we find that the most natural performance measure does not explain this phenomenon. We suggest a new, more nuanced performance measure that does.

### 2020-12-21 journal paper/content/sec-model.tex
Some edits after Guy's pass, removing some color-edits.

\item \SoftMaxRandom: $\respF$ lies in $[\eps_0,1-\eps_0]$, breaks ties fairly, and has a bounded derivative around $0$ (see Definition~\ref{def:SoftMax} for a formal definition).
\end{itemize}

\HardMaxRandom and \SoftMaxRandom can be seen as agent types that\gadelete{, resp., give some chance to the other principal and avoid sharp transitions in their probabilistic response.} \gaedit{make mistakes due to being oblivious to the principals' reputation, due to lack of awareness or interest. \HardMaxRandom models users as either being aware of the reputation or not, whereas \SoftMaxRandom models user mistakes as following a smooth density.}
\asmargincomment{I don't get YK's beef with this.}
Alternatively, they can be realized as a distribution over more ``basic" agent types. Indeed, \gaedit{the} \HardMaxRandom distribution is a mixture of \HardMax and ``random agents" (which choose a principal uniformly at random). \gadelete{The latter can be interpreted as consumers oblivious to principals' reputation, due to the lack of awareness or interest.} One can obtain a \SoftMaxRandom response function using agent types that choose a principal $i$ with a largest reward estimate $\est_i$, unless $|\est_1-\est_2|$ is upper-bounded by some parameter $\theta$, in which case they choose uniformly. Then, we obtain \SoftMaxRandom as a mixture of random agents and these ``$\theta$-\HardMax'' agents, for a suitable distribution over $\theta$.

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm. \gaedit{Principals are risk-neutral and their utility is defined as their market share, \ie the number of agents that chose this principal, and thus they aim to select the algorithm that maximizes their expected utility. (Since MAB algorithms can be randomized, we make no distinction between pure and mixed strategies.)}
\gadelete{Principal's utility is defined as the market share, \ie the number of agents that chose this principal. Principals are risk-neutral and aim to maximize their expected utility.}

>>>>

\item \SoftMaxRandom: $\respF$  lies in $[\eps_0,1-\eps_0]$, breaks ties fairly, and has a bounded derivative around $0$ (see Definition~\ref{def:SoftMax} for a formal definition). \asedit{Intuitively, $\respF$ is a smoothed version of \HardMaxRandom, without a sharp threshold therein.}
\end{itemize}

\asedit{\HardMaxRandom and \SoftMaxRandom agents can be interpreted in several ways. First, they make mistakes, due to lack of awareness or interest. Second, they give some chance to the non-preferred principal, due to irrational effects or curiosity.}
%\gaedit{make mistakes due to being oblivious to the principals' reputation, due to lack of awareness or interest. \HardMaxRandom models users as either being aware of the reputation or not, whereas \SoftMaxRandom models user mistakes as following a smooth density.}
Third, they can be realized as a distribution over more ``basic" agent types. Indeed, the \HardMaxRandom distribution is a mixture of \HardMax and ``random agents" (which choose a principal uniformly at random). \asedit{The latter can be interpreted as consumers that are completely oblivious to principals' reputation.}
%, due to the lack of awareness or interest.} 
One can obtain a \SoftMaxRandom response function using agent types that choose a principal $i$ with a largest reward estimate $\est_i$, unless $|\est_1-\est_2|$ is upper-bounded by some parameter $\theta$, in which case they choose uniformly. Then, we obtain \SoftMaxRandom as a mixture of random agents and these ``$\theta$-\HardMax'' agents, for a suitable distribution over $\theta$.

@@

\xhdr{Competition game.}
Some of our results explicitly study the game between the two principals, termed the \emph{competition game}. We model it as a simultaneous-move game: before the first agent arrives, each principal commits to an MAB algorithm. \asedit{(Since MAB algorithms can be randomized, we make no distinction between pure and mixed strategies.) Principals are risk-neutral and their utility is defined as their market share, \ie the number of agents that chose this principal. Thus they aim to select the algorithm that maximizes their expected market share.}

%Principal's utility is defined as the market share, \ie the number of agents that chose this principal. Principals are risk-neutral and aim to maximize their expected utility.

### 2020-12-21 journal paper/content/sim_details.tex
Some edits after Guy's pass, removing some color-edits.

We look for qualitative differences between these three classes under competition. We take a representative algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other representative algorithms are chosen.
We use \Thompson (\TS) from the ``adaptive-exploration" algorithms\footnote{\gaedit{We provide details on \Thompson in Appendix~\ref{app:bg}.}} \DynamicEpsGreedy (\DEG) from the ``exploration-separating" algorithms,%
\footnote{\gaedit{\DynamicEpsGreedy proceeds as follows. In each round, with probability $\eps$ it uniformly at random selects an arm to pull from the full set of arms and} ``exploits" with the remaining probability, choosing an arm with maximal posterior mean reward given the current data. We use $\eps=5\%$ throughout. Our pilot experiments show that choosing a different $\eps$ does not qualitatively change the results.}  and \DynGreedy (\DG) algorithm  as in Section~\ref{sec:theory-prelims}. For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.

>>>>

We look for qualitative differences between these three classes under competition. We take a representative algorithm from each class. Our pilot experiments indicate that our findings do not change substantially if other representative algorithms are chosen. We use 
\DynGreedy (\DG) algorithm  as in Section~\ref{sec:theory-prelims},
\DynamicEpsGreedy (\DEG) from the ``exploration-separating" algorithms,
and \Thompson (\TS) from the ``adaptive-exploration" algorithms.%
\footnote{\asedit{In each round $t$, \Thompson computes a Bayesian posterior on $\mu_a$ for each arm $a$ and draws an independent sample $\tilde{\mu}_{a,t}$ from this posterior; it chooses an arm which maximizes $\tilde{\mu}_{a,t}$.
    \newline\indent 
\DynamicEpsGreedy proceeds as follows. In each round, with probability $\eps$ it explores by choosing an arm from the full set of arms uniformly at random. With the remaining probability,} it ``exploits" by choosing an arm with maximal posterior mean reward given the current data. We use $\eps=5\%$ throughout. Our pilot experiments show that choosing a different $\eps$ does not qualitatively change the results.}
For ease of comparison, all three algorithms are parameterized with the same ``fake" Bayesian prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.
